Handling String-Number Conversion in String Constraint Solving

### 2019-11-08 draft.tex
under

\newcommand{\nat}{\mathbb{N}}
\newcommand{\todo}[1]{{\color{blue}TODO: #1}}
\newcommand{\lh}[1]{{\color{orange}Lukas: #1}}
\newcommand{\yfc}[1]{{\color{blue}YFC: #1}}
\newcommand{\petr}[1]{{\color{pink}Petr: #1}}
\newcommand{\sti}[1]{\mbox{\textsf{toInt}($#1$)}}
\newcommand{\its}[1]{\mbox{\textsf{toStr}($#1$)}}
\newcommand{\varn}{\mbox{$\mathbb{V}_{\mathbb{Z}}$}}
\newcommand{\vars}{\mbox{$\mathbb{V}_{\Sigma^*}$}}
\newcommand{\cvars}{\mbox{$\mathbb{V}_{\Sigma_\epsilon}$}}
\newcommand{\pvars}{\mbox{$\mathbb{V}_{\sharp}$}}
%\newcommand{\cvar}{\mbox{$V_{\Sigma_\epsilon}$}}
%\newcommand{\cvarone}{\mbox{$V_{\Sigma_\epsilon}$}}
%\newcommand{\cvartwo}{\mbox{$V'_{\Sigma_\epsilon}$}}
\newcommand{\cvarone}{V}
\newcommand{\cvartwo}{V'}
\newcommand{\cvar}{V}
\newcommand{\pvar}{V_\#}
\newcommand{\pvarone}{V_\#}
\newcommand{\pvartwo}{V'_\#}
\newcommand{\modelsof}[1]{[\![#1]\!]}
\newcommand{\true}{\mbox{$\mathsf{true}$}}
\newcommand{\false}{\mbox{$\mathsf{false}$}}
\newcommand{\enc}[1]{[\![#1]\!]}
%\newcommand{\parikhof}[1]{\mathit{Parikh}{(#1)}}
\newcommand{\parikhof}[1]{\mathbb{P}{(#1)}}
\newcommand{\parikhwof}[2]{|#1|_{#2}}
%\newcommand{\semof}[1]{||#1||}
\newcommand{\semof}[1]{\modelsof{#1}}
\newcommand{\decode}[1]{\mathit{decode_{#1}}}
\newcommand{\parikhfof}[1]{\Phi_{\parikhof{#1}}}
\newcommand{\pim}{\alpha}
\newcommand{\syncop}{\Cap}
\newcommand{\syncof}[2]{#1 \syncop #2}
\newcommand{\syncfof}[2]{\Psi_{\syncof {#1} {#2}}}
\newcommand{\syncT}{T_\syncop}

@@

We use $\mathbb{N}$ and $\mathbb{Z}$ to denote the sets of natural numbers (including 0) and 
integers. For a set $A$, we use $|A|$ to denote its size. 
An \emph{alphabet} is a finite set $\Sigma$ of \emph{characters} and a \emph{string} over $\Sigma$ is a sequence $w = a_1\ldots a_n$ of characters from $\Sigma$, with $\epsilon$ denoting the \emph{empty string}. 
We use $w_1\cdot w_2$ to denote the \emph{concatenation} of strings $w_1$ and $w_2$.
$\Sigma^*$ is the set of all strings over $\Sigma$, $\Sigma^+ = \Sigma^*\setminus \{\epsilon\}$, and 
A \emph{language} over $\Sigma$ is a subset $L$ of $\Sigma^*$.  
%
We use $|w|$ to denote the length of $w$ and $|w|_a$ to denote the number of occurrences of the character $a\in \Sigma$ in $w$.  
The \emph{Parikh image} of $w$ is the assignment 
%$\parikh(w) = \{a_1 \mapsto |w|_{a_1},\ldots, a_n \mapsto |w|_{a_n}\}$ that maps every symbol in $w$ to the number of its occurences in $w$. 
$\parikhof w$ that for every character $a\in\Sigma$ maps the \emph{Parikh variable} $\#a$ to the number of occurrences of $a$ in $w$, i.e., $\parikhof w (\#a) = |w|_a$. 
The Parikh image of a language $L$ is the set of Parikh images $\parikhof L = \{\parikhof w \mid w \in  L\}$.

@@

It is well known that the Parikh image of the language of $A$ can be efficiently computed in the form of linear arithmetic formula.
We will use $\parikhof{A}$ as a shorthand for $\parikhof{L(A)}$.

\begin{lemma}[\cite{PeterParikh}]
$\parikhof{A}$ of an FA $A$ can be computed in the form of a linear arithmetic formula  $\parikhfof{A}$ such that $\modelsof{\parikhfof A} = \parikhof{A}$, in time linear to the size of $A$.
\end{lemma}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Parametric Flat Automata} \label{section:sfa}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\st}[2]{q_{#1}^{#2}}
\newcommand{\sym}[2]{v_{#1}^{#2}}
\todo{add Presburger constraints to allow more expressive power.}
%We use $\cvar$ to denote a finite set of \emph{character variables}  ranging over $\Sigma_\epsilon$. 
%Sometimes we use index to denote different set of variables over the same range, e.g., $\cvarone$ and $\cvartwo$. 
%We use $v$ and its variants for character variables. We call a string over $\cvar$ (character variables) a \emph{symbolic string}.
%An \emph{interpretation $I$ over $\cvar$} is a mapping from $\cvar$ to $\Sigma_\epsilon$. For a symbolic string $x= v_1v_2\cdots v_k$ over $\cvar$, $I(x)$ is defined as $I(v_1)\cdot I(v_2)\cdot \ldots \cdot I(v_k)$.

@@

Since every string $w \in \semof{A}$ is obtained from a parametric string $x\in L(A)$ and an interpretation of character varriables $I:\cvar\rightarrow\Sigma_\epsilon$ such that $w = I(x)$, and since $x = \decode A(\parikhof x)$, 
we may say that the assignment $I \cup \parikhof x$ uniting the interpretation $I$ and the Parikh image of the parametric string $x$ encodes $w$. Lemma~\ref{lemma:parikh} also implies that two different strings are encoded by different united assignments (the Parikh image or the interpretation must be different), and hence we may extend the decoding function $\decode A$ to united assignments so that $\decode A(I \cup x) = w$.
Encodings of strings in $\semof{A}$ are encodings of parametric strings of $L(A)$ paired with arbitrary interpretations of character variables $\cvar$, 
hence $\semof A$ is characterised by models of $\parikhof {A}$ interpreted over variables $\cvar\cup\pvar$ (variables of $\cvar$  may have arbitrary values since $\parikhof {A}$ does not constraint them), we may say that 
%\begin{lemma}\ref{lemma:decw}
%For a PFA $A$, $\semof{A} = \{\decode(I\cup \pim) \mid \pim\in\semof{\parikhfof A} \land I:\cvar\rightarrow \Sigma_\epsilon\}$.
%\end{lemma}
 $$\semof{A} = \{\decode A(I\cup \pim) \mid I\cup\pim\in\semof{\parikhfof A}_{\cvar\cup\pvar}\}$$
where $\semof{\parikhfof A}_{\cvar\cup\pvar}$ are models of $\parikhfof A$, assignments to $\pvar$, united with arbitrary interpretations of $\cvar$ \lh{this is ugly}\yfc{I totally agree}.

\paragraph{Synchronisation of PFA}
The following construction of the so called synchronisation formula of two PFA $A$ and $A'$ is a crucial component of building overapproximations of string constraints. 
Synchronisation formula is a formula $\syncfof A {A'}$ over $\pvarone\cup\pvartwo\cup\cvarone\cup \cvartwo$ encoding the semantic intersection $\semof A \cap \semof {A'}$. Besides the intersection, it contains a more precise information about the synchronisation of $A$ and ${A'}$, 
namely, it characterises the set of pairs of encodings $(I\cup \pim)\in \parikhof {A}$ and $(I'\cup \pim')\in\parikhof {A'}$ such that 
$\decode A(I\cup \pim) = \decode {A'}(I'\cup \pim')$.% which constitute the intersection.

Assume that $A = (Q,T,\cvarone,q_i,q_f)$ and ${A'} = (Q',T',\cvartwo,q_i',q_f')$ are two SFA over disjoint sets of variables, $\cvarone \cap \cvartwo = \emptyset$. 
We build so called asynchronous product automaton $\syncof A {A'}$ of $A$ and ${A'}$ and then extract the formula $\syncfof A {A'}$ from its Parikh image.
%
Such product automaton uses $Q\times Q'$ as the set of states and 
%$(\cvarone \cup \{\epsilon\}) \times (\cvartwo \cup \{\epsilon\} )$ 
$\cvarone_\epsilon\times\cvartwo_\epsilon$
as the alphabet. 
%
Every accepting run of $\syncof A {A'}$ \lh{define run as a sequence of transitions etc} corresponds to an accepting run of $A$ and an accepting run of ${A'}$, and the word accepted by the run of $\syncof A {A'}$ implies a set of constraints on $\pvarone\cup\pvartwo\cup\cvarone\cup\cvartwo$ under which the word accepted by $A$ has the same semantics as the word accepted by ${A'}$.
%
Particularly, if a transition $((q_1,q_1'), (v,v'),(q_2,q_2'))$ is taken, it means the character variable $v$ and $v'$ should be assigned the same value and this value leads $A$ from state $q_1$ to state $q_2$ and ${A'}$ from $q_1'$ to $q_2'$.
%
If a transition $((q_1,q_1'), (v,\epsilon),(q_2,q_1'))$ is taken, then it means that the character variable $v$ should be assigned $\epsilon$, $A$ may transition under $v$ to $q_1$ and ${A'}$ stays in the same state, as a response to $A$s reading epsilon may as well be taking no action. Symmetrically, ${A'}$ might read a variable $v$ assigned $\epsilon$ and $A$ may stay.

Formally, the product automaton is a tuple $\syncof A {A'} = (Q\times Q', \syncT, \cvarone_\epsilon \times \cvartwo_\epsilon, (q_i,q_i'),(q_f,q_f'))$, where the transition relation $\syncT$ is the minimal set satisfying the following:
%
\begin{itemize}
\item If $(q_1,v,q_2) \in T$ and $(q_1',v',q_2') \in T'$, then we have $((q_1,q_1'),(v,v'),(q_2,q_2'))\in \syncT$.
\item If $(q_1,v,q_2) \in T$, then for all states $q'\in Q'$, we have $((q_1,q'),(v,\epsilon),(q_2,q'))\in \syncT$.
\item If $(q_1',v',q_2') \in T'$, then for all states $q\in Q$, we have $((q,q_1'),(\epsilon,v'),(q,q_2'))\in \syncT$.
\end{itemize}	
%
Next, we define the set of constraints that extract from $\syncof A {A'}$ pairs of united assignments (encodings of strings over $\Sigma$) that encode the same string. 
We start with the Parikh formula $\parikhfof {\syncof A {A'}}$ of $\syncof A {A'}$ over the set of variables $\{\#(v,v')\mid (v,v') \in \cvarone_\epsilon \times \cvartwo_\epsilon\}$ of the product automaton and conjoin it with the following restricting constraints. 
%We use $\#v$ and $\#v'$ to denote the number of occurrences of $v_i$ and $v_i^{(j)}$, respectively.

@@

The synchronisation formula is the conjunction 
$$\label{lemma:synccorrect}
\syncfof A {A'} = \parikhfof{\syncof A {A'}} \land \Psi_{\#} \land \Psi_=
$$
The correctness of this construction is expressed by Lemma~\ref{lemma:synccorrect}. Informally,
the

\begin{lemma}
 $\semof {\exists .\,\syncfof A {A'}} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
\end{lemma}

>>>>

\newcommand{\nat}{\mathbb{N}}
\newcommand{\todo}[1]{{\color{blue}TODO: #1}}
\newcommand{\lh}[1]{{\color{orange}Lukas: #1}}
\newcommand{\yfc}[1]{{\color{blue}YFC: #1}}
\newcommand{\petr}[1]{{\color{pink}Petr: #1}}
\newcommand{\sti}[1]{\mbox{\textsf{toInt}($#1$)}}
\newcommand{\its}[1]{\mbox{\textsf{toStr}($#1$)}}
\newcommand{\varn}{\mbox{$\mathbb{V}_{\mathbb{Z}}$}}
\newcommand{\vars}{\mbox{$\mathbb{V}_{\Sigma^*}$}}
\newcommand{\cvars}{\mbox{$\mathbb{V}_{\Sigma_\epsilon}$}}
\newcommand{\pvars}{\mbox{$\mathbb{V}_{\sharp}$}}
%\newcommand{\cvar}{\mbox{$V_{\Sigma_\epsilon}$}}
%\newcommand{\cvarone}{\mbox{$V_{\Sigma_\epsilon}$}}
%\newcommand{\cvartwo}{\mbox{$V'_{\Sigma_\epsilon}$}}
\newcommand{\cvarone}{V}
\newcommand{\cvartwo}{V'}
\newcommand{\cvar}{V}
\newcommand{\modelsof}[1]{[\![#1]\!]}
\newcommand{\true}{\mbox{$\mathsf{true}$}}
\newcommand{\false}{\mbox{$\mathsf{false}$}}
\newcommand{\enc}[1]{[\![#1]\!]}
%\newcommand{\parikhof}[1]{\mathit{Parikh}{(#1)}}
\newcommand{\parikhof}[1]{\mathbb{P}{(#1)}}
\newcommand{\parikhwof}[2]{|#1|_{#2}}
%\newcommand{\semof}[1]{||#1||}
\newcommand{\semof}[1]{\modelsof{#1}}
\newcommand{\decode}[1]{\mathit{decode_{#1}}}
\newcommand{\parikhfof}[1]{\Phi_{\parikhof{#1}}}
\newcommand{\pim}{\alpha}
\newcommand{\syncop}{\Cap}
\newcommand{\syncof}[2]{#1 \syncop #2}
\newcommand{\syncfof}[2]{\Psi_{\syncof {#1} {#2}}}
\newcommand{\syncT}{T_\syncop}
\newcommand{\pvarsof}[1]{\#(#1)}
\newcommand{\pvar}{\pvarsof V}
\newcommand{\pvarone}{\pvarsof V}
\newcommand{\pvartwo}{\pvarsof {V'}}

@@

We use $\mathbb{N}$ and $\mathbb{Z}$ to denote the sets of natural numbers (including 0) and 
integers. For a set $A$, we use $|A|$ to denote its size. 
An \emph{alphabet} is a finite set $\Sigma$ of \emph{characters} and a \emph{string} over $\Sigma$ is a sequence $w = a_1\ldots a_n$ of characters from $\Sigma$, with $\epsilon$ denoting the \emph{empty string}. 
We use $w_1\cdot w_2$ to denote the \emph{concatenation} of strings $w_1$ and $w_2$.
$\Sigma^*$ is the set of all strings over $\Sigma$, $\Sigma^+ = \Sigma^*\setminus \{\epsilon\}$, and 
A \emph{language} over $\Sigma$ is a subset $L$ of $\Sigma^*$.  
%
We use $|w|$ to denote the length of $w$ and $|w|_a$ to denote the number of occurrences of the character $a\in \Sigma$ in $w$.

@@

The \emph{Parikh image} of a wore $w\in \Sigma^*$ is the assignment 
$\parikhof w$ that for every character $a\in\Sigma$ maps the \emph{Parikh variable} $\#a$ to the number of occurrences of $a$ in $w$.
Formally, let $\pvarsof S$ denote the set of Parikh variables $\{\#s \mid s\in S\}$ obtained from a set $S$. 
The Parikh image of $w$ is a function $\parikhof w:\pvarsof \Sigma \rightarrow \nat$ such that $\parikhof w (\#a) = |w|_a$. 
The Parikh image of a language $L$ is the set of Parikh images $\parikhof L = \{\parikhof w \mid w \in  L\}$ and the Parikh image of an FA $A$ is Parikh image of its language, $\parikhof{A} = \parikhof{L(A)}$. 
%
It is well known that $\parikhof A$ can be efficiently computed in the form of linear interger arithmetic formula:

\begin{lemma}[\cite{SeidlSMH04}]
$\parikhof{A}$ of an FA $A$ can be computed in the form of a linear interger arithmetic formula  $\parikhfof{A}$ such that $\modelsof{\parikhfof A} = \parikhof{A}$, in time linear to the size of $A$.
\end{lemma}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Parametric Flat Automata} \label{section:sfa}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\st}[2]{q_{#1}^{#2}}
\newcommand{\sym}[2]{v_{#1}^{#2}}
\todo{add Presburger constraints to allow more expressive power.}
\lh{TODO: solve the where to add $-1$ problem}
%We use $\cvar$ to denote a finite set of \emph{character variables}  ranging over $\Sigma_\epsilon$. 
%Sometimes we use index to denote different set of variables over the same range, e.g., $\cvarone$ and $\cvartwo$. 
%We use $v$ and its variants for character variables. We call a string over $\cvar$ (character variables) a \emph{symbolic string}.
%An \emph{interpretation $I$ over $\cvar$} is a mapping from $\cvar$ to $\Sigma_\epsilon$. For a symbolic string $x= v_1v_2\cdots v_k$ over $\cvar$, $I(x)$ is defined as $I(v_1)\cdot I(v_2)\cdot \ldots \cdot I(v_k)$.

@@

Since every string $w \in \semof{A}$ is obtained from a parametric string $x\in L(A)$ and an interpretation of character varriables $I:\cvar\rightarrow\Sigma_\epsilon$ such that $w = I(x)$, and since $x = \decode A(\parikhof x)$, 
we may say that the assignment $I \cup \parikhof x$ uniting the interpretation $I$ and the Parikh image of the parametric string $x$ encodes $w$. Lemma~\ref{lemma:parikh} also implies that two different strings are encoded by different united assignments (the Parikh image or the interpretation must be different), and hence we may extend the decoding function $\decode A$ to united assignments so that $\decode A(I \cup x) = w$.
Encodings of strings in $\semof{A}$ are encodings of parametric strings of $L(A)$ paired with arbitrary interpretations of character variables $\cvar$, 
hence $\semof A$ is characterised by models of $\parikhof {A}$ interpreted over variables $\cvar\cup\pvar$ (variables of $\cvar$  may have arbitrary values since $\parikhof {A}$ does not constraint them), we may say that 
%\begin{lemma}\ref{lemma:decw}
%For a PFA $A$, $\semof{A} = \{\decode(I\cup \pim) \mid \pim\in\semof{\parikhfof A} \land I:\cvar\rightarrow \Sigma_\epsilon\}$.
%\end{lemma}
$$
\semof{A} = \{\decode A(I\cup \pim) \mid I\cup\pim\in\semof{\parikhfof A}_{\cvar\cup\pvar}\}
$$
where $\semof{\parikhfof A}_{\cvar\cup\pvar}$ are models of $\parikhfof A$, assignments to $\pvar$, united with arbitrary interpretations of $\cvar$ \lh{this is ugly}\yfc{I totally agree}.

\paragraph{Synchronisation of PFA}
The following construction of the so called synchronisation formula of two PFA $A$ and $A'$ is a crucial component of building underapproximations of string constraints. 
Synchronisation formula is a formula $\syncfof A {A'}$ over $\pvarone\cup\pvartwo\cup\cvarone\cup \cvartwo$ encoding the semantic intersection $\semof A \cap \semof {A'}$. Besides the intersection, it contains a more precise information about the synchronisation of $A$ and ${A'}$, 
namely, it characterises the set of pairs of encodings $(I\cup \pim)\in \parikhof {A}$ and $(I'\cup \pim')\in\parikhof {A'}$ such that 
$\decode A(I\cup \pim) = \decode {A'}(I'\cup \pim')$.% which constitute the intersection.

Assume that $A = (Q,T,\cvarone,q_i,q_f)$ and ${A'} = (Q',T',\cvartwo,q_i',q_f')$ are two SFA over disjoint sets of variables, $\cvarone \cap \cvartwo = \emptyset$. 
We build so called asynchronous product automaton $\syncof A {A'}$ of $A$ and ${A'}$ and then extract the formula $\syncfof A {A'}$ from its Parikh image.
%
Such product automaton uses $Q\times Q'$ as the set of states and 
%$(\cvarone \cup \{\epsilon\}) \times (\cvartwo \cup \{\epsilon\} )$ 
$\cvarone_\epsilon\times\cvartwo_\epsilon$
as the alphabet. 
%
Every accepting run of $\syncof A {A'}$ corresponds to an accepting run of $A$ and an accepting run of ${A'}$, and the word accepted by the run of $\syncof A {A'}$ implies a set of constraints on $\pvarone\cup\pvartwo\cup\cvarone\cup\cvartwo$ under which the word accepted by $A$ has the same semantics as the word accepted by ${A'}$.
%
Particularly, if a transition $((q_1,q_1'), (v,v'),(q_2,q_2'))$ is taken, it means the character variable $v$ and $v'$ should be assigned the same value and this value leads $A$ from state $q_1$ to state $q_2$ and ${A'}$ from $q_1'$ to $q_2'$.
%
If a transition $((q_1,q_1'), (v,\epsilon),(q_2,q_1'))$ is taken, then it means that the character variable $v$ should be assigned $\epsilon$, $A$ may transition under $v$ to $q_1$ and ${A'}$ stays in the same state, as a response to $A$s reading epsilon may as well be taking no action. Symmetrically, ${A'}$ might read a variable $v$ assigned $\epsilon$ and $A$ may stay.

Formally, the product automaton is a tuple $\syncof A {A'} = (Q\times Q', \syncT, \cvarone_\epsilon \times \cvartwo_\epsilon, (q_i,q_i'),(q_f,q_f'))$, where the transition relation $\syncT$ is the minimal set satisfying the following:
%
\begin{itemize}
\item If $(q_1,v,q_2) \in T$ and $(q_1',v',q_2') \in T'$, then we have $((q_1,q_1'),(v,v'),(q_2,q_2'))\in \syncT$.
\item If $(q_1,v,q_2) \in T$, then for all states $q'\in Q'$, we have $((q_1,q'),(v,\epsilon),(q_2,q'))\in \syncT$.
\item If $(q_1',v',q_2') \in T'$, then for all states $q\in Q$, we have $((q,q_1'),(\epsilon,v'),(q,q_2'))\in \syncT$.
\end{itemize}	
%
Next, we define the set of constraints that extract from $\syncof A {A'}$ pairs of united assignments (encodings of strings over $\Sigma$) that encode the same string. 
We start with computing the Parikh formula $\parikhfof {\syncof A {A'}}$ of $\syncof A {A'}$ over the set of variables $\pvarsof{\cvarone_\epsilon\times\cvartwo_\epsilon}$ and then conjoin it with the following restricting constraints. 
%We use $\#v$ and $\#v'$ to denote the number of occurrences of $v_i$ and $v_i^{(j)}$, respectively.

@@

The synchronisation formula is the conjunction 
$$\label{lemma:synccorrect}
%\syncfof A {A'} = \exists \pvarsof{\cvarone\times\cvartwo}\exists {(\cvarone\times\cvartwo)}.\,\parikhfof{\syncof A {A'}} \land \Psi_{\#} \land \Psi_=
\syncfof A {A'} = {(\cvarone\times\cvartwo)}.\,\parikhfof{\syncof A {A'}} \land \Psi_{\#} \land \Psi_=
$$
The correctness of this construction is expressed by Lemma~\ref{lemma:synccorrect}. Informally,
the

\begin{lemma}
 $\semof {\syncfof A {A'}} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
\end{lemma}
